Population Growth
========================================================

Deterministic Growth
-------------------------
Consider the simple discrete logistic growth model

```{r}
r = 1
K = 10
n0 = 1
NT = 30
time = 1:NT
```

can then simulate as

```{r}
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K)
}
```


You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(time,n,type='b',ylim=c(0,12))
```

Observation Error
-------------------------

One of the realities of science is that we rarely get to observe our system without there being some noise or uncertainty in the observations themselves.
```{r}
sd.o = 0.25
plot(time,n,type='l',ylim=c(0,12))
points(time,n+rnorm(NT,0,sd.o))
```

```{r,echo=3:4}
plot(time,n,type='l',ylim=c(0,12))
points(time,n+rnorm(NT,0,sd.o))
lines(time,n+1.96*sd.o,lty=2)
lines(time,n-1.96*sd.o,lty=2)
```



Observation error does not affect the underlying process itself. The process here is still deterministic, we're just observing it imperfectly


Parameter Uncertainty
-------------------------

We next want to account for parameter uncertainty.

To begin, we need to add parameters that describe our uncertainty in parameters

```{r}
r.sd = 0.2
K.sd = 1.0
NE = 10
```

```{r}
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}
```

```{r}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n')
for(i in 1:NE){
  lines(time,n[i,])
}
```
Next if we increase the ensemble size...

```{r,echo=1}
NE = 1000
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n')
for(i in 1:NE){
  lines(time,n[i,])
}
hist(rE)
hist(KE)
```

This is hard to visualize, so lets instead look at summary statistics

```{r}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

```{r}
n.stats = apply(n,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[5, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(n.stats[2, ], rev(n.stats[4, ]),
  n.stats[2, 1])), border = NA, col = 'blue')
lines(time,n.stats[3,],lwd=4,col="purple")
```

If K is well known...

```{r,echo=1}
K.sd = 0.01
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}
yrng = range(n)
hist(rE)
hist(KE)
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

Stochastic Population Growth
-------------------------

Lets return to the original model but now create the simplest stochastic version of the model by assuming additive noise

```{r}
n.sd = 0.25
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K) + rnorm(1,0,n.sd)
}
plot(time,n,type='l')
```

```{r,echo=1}
NE = 1000
n = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K) + rnorm(1,0,n.sd)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n')
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

Initial Conditions
------------------

```{r,echo=1:2}
n0.sd = 0.6
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n')
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

### sensitivity to initial conditions
```{r}
r=2.8
n0=0.1
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K)
}
plot(time,n,type='l')
```

```{r,echo=1:2}
n0.sd = 0.3
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n')
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
hist(n[,NT])
```

```{r,echo=1:2}
n0=9
n0.sd = 0.01
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
#plot(time,n[1,],ylim=yrng,type='n')
#for(i in 1:10){
#  lines(time,n[i,])
#}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

```{r,echo=1:2}
n0=9
n0.sd = 0.05
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
#plot(time,n[1,],ylim=yrng,type='n')
#for(i in 1:10){
#  lines(time,n[i,])
#}
n2.stats = n.stats
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(n2.stats[1, ], rev(n2.stats[3, ]),
  n2.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=3,col="green4")
lines(time,n2.stats[2,],lwd=2)
legend("bottomright",legend=c(0.05,0.01),lty=1,col=c("lightgreen","lightblue"),lwd=15,cex=1.5)
```

Rapidly diverge to 'climatology', whereby the model prediction does no better than summary statistics

### Foreshadowing: Data Assimilation in a Chaotic System
```{r}
r=2.8
K=10
n0=9.01
sd.o = 0.25
n.true = rep(n0,NT)
for(t in 2:NT){
  n.true[t] = n.true[t-1] + r*n.true[t-1]*(1-n.true[t-1]/K)
}
n.obs = rnorm(NT,n.true,sd.o)
plot(time,n.true,type='l')
points(time,n.obs,pch=18)
legend("bottomright",legend=c("True","Obs"),lty=1:0,pch=c(NA,18),lwd=2)
```

```{r,echo=FALSE}
n0.sd = 0.01
n0s = rlnorm(NE,log(n0),n0.sd)
ne = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))

## initial ensemble
ens = matrix(NA,NT*2+1,NE)
ens[1,] = rlnorm(NE,log(n0),n0.sd)
Xae = Pae = 0
Xfe = mean(ens[1,])
Pfe = var(ens[1,])
for(i in 1:NT){
  
  ##Approximate forecast with Normal
  k = 2*(i-1)+1
  Xfe[i] = mean(ens[k,])
  Pfe[i] = var(ens[k,])
  
  ## analysis step: combine previous forecast with observed data
  H <- matrix(1,ncol=1)
  Kg <- Pfe[i] * t(H) %*% solve(H%*%Pfe[i]%*%t(H) + sd.o)
  Xae[i] <- Xfe[i] + Kg*(n.obs[i] - Xfe[i]) 
  Pae[i] <- (1-Kg %*% H)*Pfe[i]    
  
  ## generate new ensemble
  ens[k+1,] = rnorm(NE,Xae[i],sqrt(Pae[i]))
  
  ## Forecast step: predict to next step from current
  ens[k+2,] = ens[k+1,] + r*ens[k+1,]*(1-ens[k+1,]/K)

#  Xfe[i+1] = m*Xa[i] + beta[2]
#  Pfe[i+1] = q + m*Pa[i]*m
  
}
fx = ens[2*(1:NT)-1,]
fx.stats = apply(fx,1,quantile,c(0.025,0.5,0.975))
plot(time,fx.stats[2,],col="blue",ylim=yrng,type='n')
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[1, ], rev(fx.stats[3, ]),
  fx.stats[1, 1])), border = NA, col = 'lightgreen')
#lines(time,fx.stats[2,],lwd=4,col="green4")
#lines(time,n.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("bottomright",legend=c("True","Obs","EnKF","Model"),lty=c(1,0,1,1),col=c(1,2,"lightgreen","lightblue"),lwd=c(1,NA,15,15),cex=1.25,pch=c(NA,18,NA,NA))
```
